An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-04-01 paper.tex
Changes suggested by Neil.
Made sure to note that Littenberg and Cornish suggested a similar
method first.

Footnoted that our kD tree is unusual in that it places box boundaries
between the points, while a standard kD tree places boundaries on the
points.

Implemented naively, RJMCMC has a significant drawback: because the
chain of samples must be Markovian, only the current sample is
available to the algorithm as it is choosing the next sample.  Each
time an RJMCMC transitions between models, the information about the
choices of parameter values in the previous model is lost; subsequent
jumps into that model must ``start fresh,'' and are correspondingly
unlikely to be accepted, delaying convergence of the RJMCMC sample
chain.

Here we introduce a technique for improving the acceptance ratio of
inter-model jumps in an RJMCMC, leading to dramatically improved
convergence of RJMCMC sample chains.  The technique uses a kD-tree
data structure to construct an approximation to each model's posterior
parameter distribution.  We draw jump proposals into the model from this approximation to its posterior.  Because jumps are proposed
preferentially to locations favored by the single-model posterior, the
RJMCMC compares ``good'' locations in parameter space across all the
models, and convergence is generally rapid.  We have successfully applied this RJMCMC technique to a 10-way model selection among alternative mass distribution models for black-hole X-ray binaries \cite{Farr2010}.

Reference \cite{Littenberg2009} proposed an alternate method for
producing inter-model jumps in an RJMCMC that relies on a box
decomposition of parameter space, but uses fixed-sized boxes.  The box
size is fixed, so the method cannot adapt to the local structure of
the posterior, and becomes asymptotically inefficient for
high-dimensional parameter spaces or highly peaked posteriors.
Meanwhile, the approximation to the posterior distribution produced by
the kD-tree is a constant-in-box interpolation of the posterior,
similar in spirit to the phase-space density interpolants produced
from N-body positions and momenta in Reference \cite{Ascasibar2005}.
The kD-tree interpolation is effective in parameter spaces of
arbitrary dimensionality, and is quite space-efficient, requiring
$\order{N}$ storage space and $\order{\log N}$ time to produce each
proposed jump, where $N$ is the number of single-model MCMC samples
used to construct the interpolation.

@@

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively:
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to Step 1 with each of the sub-boxes, storing the
  resulting trees as sub-trees of the current box.
\end{enumerate}
The key algorithmic step in the production of a kD-tree is finding the
median point along a given dimension in order to divide the points in
half in Step 2.  For $n$ points, this can be accomplished in
$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
points in total, there are $\order{\log N}$ levels in the tree; at
each level, $\order{N}$ points must be processed once in the
median-finding algorithm.  Tree construction thus costs $\order{N \log
  N}$ in time, and the tree consumes $\order{N}$ space.  As an
illustration, box boundaries for a kD-tree constructed around a point
set that is normally distributed around the origin in two dimensions
are shown in Figure \ref{fig:kD-tree}.

>>>>

Implemented naively, RJMCMC has a significant drawback: because the
chain of samples must be Markovian, only the current sample is
available to the algorithm as it is choosing the next sample.  Each
time an RJMCMC transitions between models, the information about the
choices of parameter values in the previous model is lost; subsequent
jumps into that model must ``start fresh,'' and are correspondingly
unlikely to be accepted, delaying convergence of the RJMCMC sample
chain.  One of the first papers to address this issue was Reference
\cite{Littenberg2009}, where the authors propose a method for
producing inter-model jumps in an RJMCMC that relies on interpolating
single-model posterior distributions using a box decomposition of
parameter space.

Here we introduce an alternative technique for improving the
acceptance ratio of inter-model jumps in an RJMCMC, leading to
dramatically improved convergence of RJMCMC sample chains.  The
technique uses a kD-tree data structure to construct an approximation
to each model's posterior parameter distribution.  We draw jump
proposals into the model from this approximation to its posterior.
Because jumps are proposed preferentially to locations favored by the
single-model posterior, the RJMCMC compares ``good'' locations in
parameter space across all the models, and convergence is generally
rapid.  We have successfully applied this RJMCMC technique to a 10-way
model selection among alternative mass distribution models for
black-hole X-ray binaries \cite{Farr2010}.

The method of Reference \cite{Littenberg2009} for producing
inter-model jumps in an RJMCMC relies on a box decomposition of
parameter space, using fixed-sized boxes.  The method cannot adapt to
the local structure of the posterior, and becomes asymptotically
inefficient for high-dimensional parameter spaces or highly peaked
posteriors.  Meanwhile, the approximation to the posterior
distribution produced by the kD-tree is a constant-in-box
interpolation of the posterior, similar in spirit to the phase-space
density interpolants produced from N-body positions and momenta in
Reference \cite{Ascasibar2005}.  The kD-tree interpolation is
effective in parameter spaces of arbitrary dimensionality, and is
quite space-efficient, requiring $\order{N}$ storage space and
$\order{\log N}$ time to produce each proposed jump, where $N$ is the
number of single-model MCMC samples used to construct the
interpolation.

@@

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively%
\footnote{The kD-tree datastructure defined here places box boundaries
  between the points.  An alternate definition common in the
  literature places box boundaries on the median point, but such a
  definition is inconvenient for our purposes.}: %
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to Step 1 with each of the sub-boxes, storing the
  resulting trees as sub-trees of the current box.
\end{enumerate}
The key algorithmic step in the production of a kD-tree is finding the
median point along a given dimension in order to divide the points in
half in Step 2.  For $n$ points, this can be accomplished in
$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
points in total, there are $\order{\log N}$ levels in the tree; at
each level, $\order{N}$ points must be processed once in the
median-finding algorithm.  Tree construction thus costs $\order{N \log
  N}$ in time, and the tree consumes $\order{N}$ space.  As an
illustration, box boundaries for a kD-tree constructed around a point
set that is normally distributed around the origin in two dimensions
are shown in Figure \ref{fig:kD-tree}.
